---
title: "PROBLEM SET 4"
author: "Deeptarka Saha 732"
date: "2026-02-19"
output: html_document
---

# PROBLEM SET 4


## Problem 1: Problem to demonstrate multicollinearity
```{r}
rm(list=ls())
library(ISLR)
attach(Credit)
head(Credit)
df=Credit[,c(3,4,6,12)]
head(df)
```
```{r}
#(a)
par(mfrow=c(1,2))
plot(Limit,Rating,main="Scatterplot of Rating vs Limit")
plot(Limit,Age,main="Scatterplot of Age vs Limit")
par(mfrow=c(1,1))
```
Comment:

Age vs Limit:

1. Points are widely scattered.

2. Age does not appear strongly related to credit limit.


Rating vs Limit:

1. Strong positive linear relationship.

2. As Rating increases, Limit increases almost perfectly linearly.

```{r}
#(b)
fit1=lm(Balance~Age+Limit)
fit2=lm(Balance~Rating+Age+Limit)
fit3=lm(Balance~Rating+Limit)
library(stargazer)
stargazer(fit1,fit2,fit3,type="text",out="f2.txt")
```
Observed Marked Differences :

The major difference appears in Model (ii):

In Model (i), Limit is highly significant

In Model (iii), Rating is highly significant

In Model (ii)** (when both Rating and Limit are included) one of them becomes insignificant or changes sign

```{r}
#(c) VIF
library(car)
vif(fit1)
vif(fit2)
vif(fit3)
```
We  observe:
Very high VIF for Limit and Rating ,while low VIF for Age. Hence, strong multicollinearity exists between them.

## Problem 2 :
```{r}
rm(list=ls())
library(MASS)
head(Boston)
dim(Boston)
df=Boston
df.cont=Boston
df.cont$chas = NULL
df.cont$rad = NULL
model=lm(medv~.,data=df.cont)
summary(model)
residualPlot(model)
std.resid=rstandard(model)
```
We observe :
A few observations have large studentized residuals → response outliers

Some observations have high leverage (extreme crim or lstat values) using hat- values.

A small number exceed Cook’s distance threshold → influential points