---
title: "Predictive Analytics Problem Set 4"
author: "Deeptarka Saha"
date: "2026-03-29"
output: html_document
---

## 1. Problem to demonstrate the fitting of a binary response variable using logistic regression.

```{r}
# a) Test- Train split
library(ISLR)

data(Weekly)

# Training: 1990–2008
train = Weekly[Weekly$Year <= 2008, ]

# Test: 2009–2010
test = Weekly[Weekly$Year > 2008, ]

train
test
```
```{r}
# b) Logistic Regression Model
fit = glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume,
          data = train, family = binomial)

summary(fit)
```
We observe that, typically no predictors are stongly significant at 5% level.
Usually Lag 2 test has the smallest p value.

Interpretation of  coefficient corresponding to Lag 2 test :
 
 If coefficient  of Lag2 ≈ 0.01(say):
 
 A 1 unit increase in Lag2 increases log-odds of “Up” by 0.05

Odds increase by:
𝑒^0.01≈1.010 

```{r}
# c) Odds of market going “up” at median values of the predictors.
medians = apply(train[, -9], 2, median)  # exclude Direction

log_odds = predict(fit, newdata = as.data.frame(t(medians)), type = "link")

odds = exp(log_odds)
odds
```
here, Odds = P(Up)/P(Down)

```{r}
# e) Optimal Cutoff & Test Performance.
library(pROC)
prob_test = predict(fit, newdata = test, type = "response") # probabilities on test data
prob_test

# Finding Optimal Cutoff
roc_test = roc(test$Direction, prob_test)

best=coords(roc_test, "best", ret = "threshold", best.method = "youden");
c_star=best[1]
```
```{r}
# Predictions
prob_test = predict(fit, newdata = test, type = "response")

pred_class = ifelse(prob_test > c_star, "Up", "Down")
pred_class = factor(pred_class, levels = c("Down", "Up"))

# Checking lengths
length(pred_class)
length(test$Direction)

# Confusion matrix
str(prob_test)
str(pred_class)
str(test$Direction)

mean(pred_class != test$Direction) # Test Error Rate
```
```{r}
# f) Model with Lag 1 and Lag 2 only
fit2 = glm(Direction ~ Lag1 + Lag2,
           data = train, family = binomial)

prob_test2 = predict(fit2, newdata = test, type = "response")

pred2 = ifelse(prob_test2 > 0.5, "Up", "Down")

mean(pred2 != test$Direction)
```
Using only Lag1 & Lag2 gives comparable performance.

Here Test error is almost similar or slightly better.

Simple models performs well.
