---
title: "Predictive  Analysis Problem Set 6"
author: "Deeptarka Saha"
date: "2026-03-25"
output: html_document
---
```{r}
rm(list=ls())
library(ISLR)
head(Hitters)
dim(Hitters)
?Hitters
```
```{r}
# 1. Data Cleaning
df=na.omit(Hitters) #1. Removing the missing rows
head(df)
dim(df)
```
```{r}
#2. Splitting into Train and Test
set.seed(123)
train.id=sample(1:263,210)
train.data=df[train.id,] #Training data
head(train.data)
test.data=df[-train.id,] #Test data
head(test.data)
dim(train.data);dim(test.data)
```

```{r}
#3. Best subset selection
library(leaps)
best.subset=regsubsets(Salary~.,data=train.data,nvmax=19)
summary(best.subset)
#Adjusted R squared*/*
adjr2=summary(best.subset)$adjr2
adjr2
cp=summary(best.subset)$cp
cp
bic=summary(best.subset)$bic
bic
coef(best.subset, which.min(bic)) #Best BIC Model
coef(best.subset, which.min(cp)) #Best cp Model
coef(best.subset, which.max(adjr2)) #Best Adjusted R squared Model
```
We fit models with 1 to 19 predictors. Then we choose best model using:
Adjusted R² → maximum
BIC → minimum
Cp → minimum
```{r}
# 4. Train MSE and Test MSE
m.cp=lm(Salary~AtBat+Hits+Walks+CAtBat+CRuns+CRBI+CWalks+Division+PutOuts+Assists+NewLeague,data=train.data)
m.bic=lm(Salary~Walks+CAtBat+CHits+CHmRun+Division,data=train.data)
m.adjr=lm(Salary~AtBat+Hits+Walks+CAtBat+CRuns+CRBI+CWalks+Division+PutOuts+Assists+Errors+NewLeague,data=train.data)
mse.train.cp=1/210*(sum((train.data$Salary-predict(m.cp))^2)) #Training MSE for cp
mse.train.bic=1/210*(sum((train.data$Salary-predict(m.bic))^2)) #Training MSE for bic
mse.train.adjr=1/210*(sum((train.data$Salary-predict(m.adjr))^2)) #Training MSE for adjR2
mse.train.bic
mse.train.adjr
mse.train.cp
y.hat.test.cp=predict(m.cp, newdata=test.data)
y.hat.test.bic=predict(m.bic, newdata=test.data)
y.hat.test.adjr=predict(m.adjr, newdata=test.data)
mse.test.cp=mean((test.data$Salary-y.hat.test.cp)^2)
mse.test.cp
mse.test.bic=mean((test.data$Salary-y.hat.test.bic)^2)
mse.test.bic
mse.test.adjr=mean((test.data$Salary-y.hat.test.adjr)^2)
mse.test.adjr
out.best=data.frame(Measure=c("Cp","BIC","AdjR2"),Training_MSE=c(mse.train.cp,mse.train.bic,mse.train.adjr),
                    Test_MSE=c(mse.test.cp,mse.test.bic,mse.test.adjr))
out.best

```
In all cases we observe:

Training MSE < Test MSE
This is expected because model is trained on training data

Observations on Model Comparison:

Model with lowest Training MSE → Best fit on training data
Model with lowest Test MSE → Best for prediction (preferred)
```{r}
# 5. Forward Selection
forward.subset=regsubsets(Salary~.,data=train.data,nvmax=19,method="forward")
summary(forward.subset)
coef(forward.subset, which.min(summary(forward.subset)$bic)) #Best BIC Model
coef(forward.subset, which.min(summary(forward.subset)$cp)) #Best cp Model
coef(forward.subset, which.max(summary(forward.subset)$adjr2)) #Best Adjusted R squared Model

m.cp3=lm(Salary~Walks+CAtBat+CHits+CRuns+CRBI+CWalks+Division+PutOuts+Assists+Errors+NewLeague,data=train.data)
m.bic3=lm(Salary~Walks+CAtBat+CHits+CRuns++CRBI+CWalks+Division,data=train.data)
m.adjr3=lm(Salary~AtBat+Hits+Walks+Years+CAtBat++CHits+CRuns+CRBI+CWalks+Division+PutOuts+Assists+Errors+NewLeague,data=train.data)

mse.train.cp3=1/210*(sum((train.data$Salary-predict(m.cp3))^2)) #Training MSE for cp
mse.train.bic3=1/210*(sum((train.data$Salary-predict(m.bic3))^2)) #Training MSE for bic
mse.train.adjr3=1/210*(sum((train.data$Salary-predict(m.adjr3))^2)) #Training MSE for adjR2

y.hat.test.cp3=predict(m.cp3, newdata=test.data)
y.hat.test.bic3=predict(m.bic3, newdata=test.data)
y.hat.test.adjr3=predict(m.adjr3, newdata=test.data)
mse.test.cp3=mean((test.data$Salary-y.hat.test.cp3)^2)
mse.test.bic3=mean((test.data$Salary-y.hat.test.bic3)^2)
mse.test.adjr3=mean((test.data$Salary-y.hat.test.adjr3)^2)

out.forward=data.frame(Measure=c("Cp","BIC","AdjR2"),Training_MSE=c(mse.train.cp3,mse.train.bic3,mse.train.adjr3),
                        Test_MSE=c(mse.test.cp3,mse.test.bic3,mse.test.adjr3))
out.forward

final=list(Best_Subset=out.best,Forward_Subset=out.forward,Backward_Subset=out.forward)
final
```
```{r}
# 6. Backward Selection
backward.subset=regsubsets(Salary~.,data=train.data,nvmax=19,method="backward")
summary(backward.subset)
coef(backward.subset, which.min(summary(backward.subset)$bic)) #Best BIC Model
coef(backward.subset, which.min(summary(backward.subset)$cp)) #Best cp Model
coef(backward.subset, which.max(summary(backward.subset)$adjr2)) #Best Adjusted R squared Model

m.cp2=lm(Salary~AtBat+Hits+Walks+CAtBat+CHits+CHmRun+CRuns+CWalks+Division+PutOuts+Assists+Errors+NewLeague,data=train.data)
m.bic2=lm(Salary~Walks+CAtBat+CHits+CHmRun+Division,data=train.data)
m.adjr2=lm(Salary~AtBat+Hits+Walks+CAtBat+CHits+CHmRun+CRuns+CWalks+Division+PutOuts+Assists+Errors+NewLeague,data=train.data)

mse.train.cp2=1/210*(sum((train.data$Salary-predict(m.cp2))^2)) #Training MSE for cp
mse.train.bic2=1/210*(sum((train.data$Salary-predict(m.bic2))^2)) #Training MSE for bic
mse.train.adjr2=1/210*(sum((train.data$Salary-predict(m.adjr2))^2)) #Training MSE for adjR2

y.hat.test.cp2=predict(m.cp2, newdata=test.data)
y.hat.test.bic2=predict(m.bic2, newdata=test.data)
y.hat.test.adjr2=predict(m.adjr2, newdata=test.data)
mse.test.cp2=mean((test.data$Salary-y.hat.test.cp2)^2)
mse.test.bic2=mean((test.data$Salary-y.hat.test.bic2)^2)
mse.test.adjr2=mean((test.data$Salary-y.hat.test.adjr2)^2)

out.backward=data.frame(Measure=c("Cp","BIC","AdjR2"),Training_MSE=c(mse.train.cp2,mse.train.bic2,mse.train.adjr2),
                    Test_MSE=c(mse.test.cp2,mse.test.bic2,mse.test.adjr2))
out.backward

```

7. Comparison of Best Subset, Forward and Backward Selection

1. In terms of Model Selection
  
  Best Subset Selection examines all possible models and gives the most optimal model for each criterion (Adj R², BIC, Cp) as :
 
 Forward Selection :
 
1. Starts with no variables and adds step-by-step.

2. May miss the best combination of predictors.

Backward Selection :

1.Starts with full model and removes variables.

2.Works well but depends on initial full model.


2.  In terms of Training MSE:

Best Subset → Lowest Training MSE (best fit)

Forward & Backward → Slightly higher Training MSE


3. In terms of Test MSE:

BIC-based models (often from any method) → Lower Test MSE

Best Subset (Adj R²) → May have higher Test MSE (overfitting)

Forward/Backward → Comparable but sometimes slightly worse