---
title: "MDTS4214 PROBLEM SET 7"
author: "Deeptarka Saha"
date: "2026-03-29"
output: html_document
---
# Comparison of least squares, ridge regression, LASSO and Elastic Net
```{r}
rm(list=ls())
library(ISLR)
df=na.omit(Hitters) #1. Removing the missing rows
dim(df)
#2. Splitting into Train and Test
set.seed(123)
test.id=sample(1:263,100)
train.data=df[-test.id,] #Training data
head(train.data)
test.data=df[test.id,] #Test data
head(test.data)
dim(train.data);dim(test.data)
#Train the model using least squares and report the train MSE and test MSE.
m1=lm(Salary~.,data=train.data)
mse.train=1/163*(sum((train.data$Salary-predict(m1))^2))
y.hat.test=predict(m1,newdata=test.data)
mse.test=mean((test.data$Salary-y.hat.test)^2)
mse.train
mse.test

#3. Fit Ridge Regression
library(glmnet)
lambda.grid = 10^seq(10, -2, length = 100)
x=model.matrix(Salary~.,train.data)[,-1]
x.test=model.matrix(Salary~.,test.data)[,-1]
head(x)
ridge.model=glmnet(x, train.data$Salary, alpha = 0,lambda=lambda.grid)
dim(coef(ridge.model)) # (i) Dimension of coefficient matrix

beta50 = coef(ridge.model)[,50]
beta60 = coef(ridge.model)[,60]

l2.50 = sqrt(sum(beta50^2))
l2.60 = sqrt(sum(beta60^2))

l2.50
l2.60 # (ii) L2 Norms
# Interpretation :
# As λ increases → coefficients shrink → L2 norm decreases
# So, l2.60 < l2.50

pred50 = predict(ridge.model, s = lambda.grid[50], newx = x.test)
pred60 = predict(ridge.model, s = lambda.grid[60], newx = x.test)

mse50 = mean((test.data$Salary - pred50)^2)
mse60 = mean((test.data$Salary - pred60)^2)

mse50
mse60 # (iii) Test MSE for λ

```
4. Plot of Coefficients vs λ
```{r}
plot(ridge.model, xvar="lambda", label=TRUE)

```
Interpretation:
As λ increases → coefficients shrink toward 0.

Some shrink faster than others.

No coefficient becomes exactly zero.

5. Plot vs L2 Norm Ratio
```{r}
plot(ridge.model, xvar="norm", label=TRUE)
```
Interpretation :

The graph becomes more intuitive than λ.

It Starts from:
1 → least squares

towards 0 → heavy shrinkage

Shape remains simlar, just rescaled.

6.Ridge with cross validation
```{r}
set.seed(123)
cv.ridge=cv.glmnet(x, train.data$Salary, alpha=0)

best.lambda=cv.ridge$lambda.min
best.lambda

pred=cv.predict=cv.predict <- predict(cv.ridge, s=best.lambda, newx=x.test)

mse.cv=mean((test.data$Salary - pred)^2)
mse.cv
```
7. LASSO Regression
```{r}
set.seed(123)
cv.lasso=cv.glmnet(x, train.data$Salary, alpha=1)

best.lambda.lasso=cv.lasso$lambda.min

lasso.pred=predict(cv.lasso, s=best.lambda.lasso, newx=x.test)

mse.lasso=mean((test.data$Salary - lasso.pred)^2)
mse.lasso

coef.lasso=coef(cv.lasso, s=best.lambda.lasso)

sum(coef.lasso!=0) # No. of non-zero coefficients.
```
8. Elastic Net
```{r}
set.seed(123)
cv.enet=cv.glmnet(x, train.data$Salary, alpha=0.5)

best.lambda.enet=cv.enet$lambda.min

enet.pred=predict(cv.enet, s=best.lambda.enet, newx=x.test)

mse.enet=mean((test.data$Salary - enet.pred)^2)
mse.enet

coef.enet=coef(cv.enet, s=best.lambda.enet)
coef.enet[coef.enet!=0] # Non-zero Coefficients
sum(coef.enet!=0)
```
9. Comparison of Key features of the Test mse of different Regression techniques:

For, Least Squares →  No regularization

 Ridge	 →  	Shrinkage
 
 LASSO	 →  	Variable selection
 
 Elastic Net	 →  	Hybrid
 
 Conclusion:
 
 Ridge reduces variance keeping all predictors.
 
LASSO removes irrelevant predictors.

Elastic Net combines both advantages of Ridge and LASSO.

Cross-validation gives optimal λ.

So,Regularized models predict the Salary of a baseball player more accurately than least squares.
